---
title: "Bayesian Analysis for WSEI"
format: pdf
editor: visual
---

```{r}
#| label: install-packages
#| eval: false
# Run this chunk once to install required R packages (if not already installed).
# Requires R >= 4.2 and a working C++ toolchain for brms/Stan.
install.packages(c(
  "tidyverse", "lubridate", "brms", "bayesplot",
  "posterior", "tidybayes", "mgcv", "janitor"
))
```

```{r}
library(tidyverse)
library(lubridate)
library(brms)
library(bayesplot)
library(posterior)
library(tidybayes)
library(mgcv)
library(janitor)
```

### 1. Data

```{r}
data_path <- "../../data/wildfire/merged_with_wsei.csv"
df_raw <- readr::read_csv(data_path, show_col_types = FALSE)

# Standardize names a bit (avoid spaces in formulas)
df <- df_raw %>%
  rename(
    station_id = `Station ID`,
    date = Date,
    lat = Latitude,
    lon = Longitude,
    aqi = AQI
  ) %>%
  mutate(
    date = as.Date(date),
    dow  = wday(date, label = TRUE),
    month = month(date),
    doy = yday(date),
    year = year(date)
  )

# glimpse(df)
# summary(df$aqi)

```

*Interpretation:* this sets up the panel structure needed for hierarchical modelling (i.e. many days per station) and ensures the model can distinguish from day-to-day variation vs. station-to-station baseline differences.

#### 1.1 Outcome choice

For **RQ1**, we model next-day AQI. That aligns with your “warning” goal and avoids same-day simultaneity.

```{r}
df <- df %>%
  arrange(station_id, date) %>%
  group_by(station_id) %>%
  mutate(aqi_next = lead(aqi, 1)) %>%
  ungroup() %>%
  filter(!is.na(aqi_next))
```

*Interpretation:* this avoids same-day simultaneity (smoke today vs AQI today can be contemporaneous).

## 2. RQ1 - Bayesian hierarchical regression

Model specification

-   Response: `aqi_next`

-   Fixed effects: wildfire lags (k0–k3), meteorology, traffic, weekday

-   Seasonality: smooth function of day-of-year (cyclic)

-   Random intercept: station-level baseline differences

You can choose one wildfire intensity family as “primary” (recommended: **hfi**) and treat frp/tfc as sensitivity checks.

```{r}
# Choose wildfire variables (primary)
wsei_vars <- c("wsei_hfi_k0","wsei_hfi_k1","wsei_hfi_k2","wsei_hfi_k3")

# Basic covariates (edit based on what exists in your file)
covars <- c("MeanTemp","TotalPrecip","MaxTemp","MinTemp","TempRange","TotalTraffic","wind_dir_deg")

# Keep only columns we need and drop missing
df_m <- df %>%
  select(station_id, date, aqi_next, dow, doy, month, year, all_of(wsei_vars), all_of(covars)) %>%
  drop_na()

# Standardize continuous predictors for stable priors
scale_cols <- c(wsei_vars, covars)
df_m <- df_m %>%
  mutate(across(all_of(scale_cols), ~ as.numeric(scale(.x)), .names = "{.col}_z"))
```

cyclic seasonality under brms formula

```{r}
f_rq1 <- bf(
  aqi_next ~
    wsei_hfi_k0_z + wsei_hfi_k1_z + wsei_hfi_k2_z + wsei_hfi_k3_z +
    MeanTemp_z + TotalPrecip_z + MaxTemp_z + MinTemp_z + TempRange_z +
    TotalTraffic_z + wind_dir_deg_z +
    dow +
    s(doy, bs = "cc") +
    (1 | station_id)
)

# Priors (work well when predictors are standardized)
priors <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 3), class = "Intercept"),
  prior(exponential(1), class = "sd"),      # station random intercept sd
  prior(exponential(1), class = "sigma")    # residual sd
)
```

Fit with MCMC

```{r}
fit_rq1 <- brm(
  formula = f_rq1,
  data = df_m,
  family = gaussian(),
  prior = priors,
  chains = 4, cores = 4, iter = 2000, warmup = 1000,
  seed = 946
)

fit_rq1
```

**Model**

ou fit a **Gaussian hierarchical regression** for next-day AQI with:

-   Wildfire exposure lags: `wsei_hfi_k0` … `wsei_hfi_k3`

-   Weather/traffic controls

-   Weekday factor

-   Cyclic seasonality spline `s(doy, bs="cc")`

-   Station random intercept `(1 | station_id)`

Data size: 9885 observations across 19 stations

**Key result: wildfire lags (the main story for RQ1)**

From the regression coefficients table:

-   **k0 (same-day exposure): +1.20 AQI** with 95% CrI **\[0.74, 1.64\]**\

    → strong positive association.

-   **k1: −0.67 AQI** with CrI **\[−1.27, −0.04\]**\

    → small negative association (barely below zero)

-   **k2: +0.30 AQI** with CrI **\[−0.36, 1.00\]**\

    → uncertain (crosses zero)

-   **k3: +4.12 AQI** with CrI **\[3.58, 4.65\]**\

    → *very strong* and highly certain positive association.\

    **How to explain it simply:** After controlling for weather, seasonality, weekday, traffic, and station differences, the model finds wildfire exposure is associated with higher next-day AQI, and the strongest association appears around **3 days later**.

**Controls: what’s plausible vs what to be careful about**

-   **TotalPrecip: −1.77** (CrI \[−2.08, −1.47\]) → precipitation lowers AQI (physically plausible).

-   **TempRange: +1.94** (CrI \[1.32, 2.57\]) → larger temp swings associated with higher AQI (plausible via boundary-layer/dispersion dynamics).

-   **TotalTraffic: −1.27** (CrI \[−1.97, −0.58\]) → statistically strong but **not a causal claim**; could reflect confounding/measurement mismatch.

-   **wind_dir_deg** shows a strong negative coefficient, but wind direction is circular; using degrees linearly can be misleading (good “future refinement” item).

    **Hierarchical part (station effects)**

<!-- -->

-   **sd(Intercept) ≈ 20.55** (CrI \[15.71, 26.21\]) → stations have very different baseline AQI levels; hierarchical structure is warranted.

-   **sigma ≈ 14.56** → remaining unexplained day-to-day variation in AQI_next.

### 3. Assumption Check

#### 3.1 Convergence

```{r}
summary(fit_rq1)$fixed %>% as.data.frame()

# R-hat / ESS overview
posterior::summarise_draws(as_draws_df(fit_rq1)) %>%
  select(variable, rhat, ess_bulk, ess_tail) %>%
  filter(str_detect(variable, "^b_")) %>%
  arrange(desc(rhat)) %>%
  head(15)

# Trace the plot
mcmc_trace(as_draws_array(fit_rq1), pars = c(
  "b_wsei_hfi_k0_z","b_wsei_hfi_k1_z","b_wsei_hfi_k2_z","b_wsei_hfi_k3_z"
))
```

R-hat and ESS; they are essentially \~1.00 with large ESS for key coefficients.
Trace plot for wildfire coefficients looks stable (chains mix well). Analysis**Interpretation:** the MCMC sampling is reliable enough to trust the posterior summaries.

#### 3.2 Posterior Predictive Check

```{r}
pp_check(fit_rq1, ndraws = 200)
pp_check(fit_rq1, type = "dens_overlay", ndraws = 100)
```

The overlay plots show the model captures the bulk distribution, but tails aren’t perfect (you can see mismatch in the far right tail in the overlay).

**Interpretation:** Gaussian errors are “okay for mean behavior,” but if extreme AQI accuracy matters, consider Student-t likelihood or tail modeling later.

#### 3.3 Nonlinearity Check

```{r}
res <- residuals(fit_rq1)
fit <- fitted(fit_rq1)[, "Estimate"]

diag_df <- df_m %>%
  mutate(resid = res[, "Estimate"], fitted = fit)

ggplot(diag_df, aes(fitted, resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Residuals vs Fitted (check nonlinearity / heteroskedasticity)",
       x = "Fitted AQI_next", y = "Residual")
```

Residuals vs fitted plot shows variance widening at higher fitted AQI (fanning).

**Interpretation:** errors are larger on bad-air days; uncertainty is not constant.

Our model is directionally right, but uncertainty grows in high-AQI regimes—consistent with CS1’s ‘extreme days are harder’ limitation.

#### 3.4 Autocorrelation

```{r}
daily_res <- diag_df %>%
  group_by(date) %>%
  summarise(mean_resid = mean(resid), .groups = "drop") %>%
  arrange(date)

acf(daily_res$mean_resid, main = "ACF of daily mean residuals")
```

The ACF of daily mean residuals shows residual structure at short lags. **Interpretation:** there’s still time dependence not fully captured. If forecasting is the priority, adding lagged AQI or AR structure could improve predictive performance.

### 4. Wildfire Lag Effects

#### 4.1 Posterior intervals for lag coefficients

```{r}
coef_df <- as_draws_df(fit_rq1) %>%
  select(starts_with("b_wsei_hfi_k")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "draw") %>%
  mutate(lag = str_extract(term, "k[0-3]"))

sum_df <- coef_df %>%
  group_by(lag) %>%
  summarise(
    mean = mean(draw),
    lo = quantile(draw, 0.025),
    hi = quantile(draw, 0.975),
    p_pos = mean(draw > 0),
    .groups = "drop"
  )

ggplot(sum_df, aes(x = lag, y = mean, ymin = lo, ymax = hi)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Wildfire (HFI) lag effects on next-day AQI (standardized)",
       y = "Posterior mean (95% CrI)", x = "Lag")
```

This plot is the “money plot” for RQ1 — it visualizes the lag coefficients with 95% credible intervals. It confirms:

-   strong positive at k0

-   uncertain at k2

-   dominant positive at k3

#### 4.2 Cumulative short-lag effect (0–3 days)

```{r}
draws <- as_draws_df(fit_rq1)
cum <- with(draws,
            b_wsei_hfi_k0_z + b_wsei_hfi_k1_z + b_wsei_hfi_k2_z + b_wsei_hfi_k3_z)

tibble(cum = cum) %>%
  summarise(
    mean = mean(cum),
    lo = quantile(cum, 0.025),
    hi = quantile(cum, 0.975),
    p_pos = mean(cum > 0)
  )
```

You computed the sum of k0–k3 posterior draws **mean 4.95**, CrI **\[4.63, 5.27\]**, and **p_pos = 1**

**Interpretation:** across the 0–3 day window, the total wildfire association with next-day AQI is almost certainly positive and large. This is very clean to communicate.

### 5. Operational Wildfire Signal

We tried to forecast the probability of high-AQI next day.

According to Air Quality Ontario, **Special Air Quality Statement (SAQS)** will be issured when the AQHI is forecast to reach or exceed **7 (High Risk)**.

```{r}
# High AQI next day = 1 if next-day AQI is in the top 10% of days in the dataset
q90 <- quantile(df_m$aqi_next, probs = 0.90, na.rm = TRUE)
q95 <- quantile(df_m$aqi_next, probs = 0.95, na.rm = TRUE)  # optional sensitivity

df_m <- df_m %>%
  mutate(
    high_next_90 = as.integer(aqi_next >= q90),
    high_next_95 = as.integer(aqi_next >= q95)
  )

```

Fit a Bayesian hierarchical logistic model

```{r}
f_rq3_90 <- bf(
  high_next_90 ~
    wsei_hfi_k0_z + wsei_hfi_k1_z + wsei_hfi_k2_z + wsei_hfi_k3_z +
    MeanTemp_z + TotalPrecip_z + MaxTemp_z + MinTemp_z + TempRange_z +
    TotalTraffic_z + wind_dir_deg_z +
    dow +
    s(doy, bs = "cc") +
    (1 | station_id)
)

fit_rq3_90 <- brm(
  formula = f_rq3_90,
  data = df_m,
  family = bernoulli(link = "logit"),
  prior = c(
    prior(normal(0, 1), class = "b"),
    prior(normal(0, 2), class = "Intercept"),
    prior(exponential(1), class = "sd")
  ),
  chains = 4, cores = 4, iter = 2000, warmup = 1000,
  control = list(adapt_delta = 0.95),   # helps avoid divergences
  seed = 946
)

fit_rq3_90
```

Logistic hierarchical model - fit Bernoulli likelihood with logit link

-   same covariates + station random intercept + seasonality spline

These are **log-odds** effects (per 1 SD increase), not AQI units.

-   **k3: 0.45** CrI \[0.32, 0.59\] → strong positive predictor of extreme-day risk Analysis\

    Odds ratio ≈ exp(0.45)=1.57 → \~57% higher odds per +1 SD.

-   **k1: 0.17** CrI \[0.05, 0.29\] → smaller positive effect

-   **k2: −0.17** CrI \[−0.31, −0.03\] → small negative (possible lag tradeoff / multicollinearity)

-   **k0 crosses 0** (not strong evidence)

    **How to explain to stakeholders:** The model’s wildfire exposure features, especially the 3-day lag, meaningfully increase the probability of a high-AQI day tomorrow.

#### 5.1 Risk score and calibration

Compute posterior mean risk and do a simple calibration curve by deciles.

```{r}
# posterior expected probability
p_hat <- fitted(fit_rq3_90, scale = "response")[, "Estimate"]
df_risk <- df_m %>% mutate(p_risk = p_hat)

cal <- df_risk %>%
  mutate(bin = ntile(p_risk, 10)) %>%
  group_by(bin) %>%
  summarise(
    mean_pred = mean(p_risk),
    obs_rate = mean(high_next_90),
    n = n(),
    .groups = "drop"
  )

ggplot(cal, aes(mean_pred, obs_rate, size = n)) +
  geom_point(alpha = 0.85) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(
    title = "Calibration: predicted risk vs observed frequency (top-decile AQI event)",
    x = "Mean predicted P(high AQI next day)",
    y = "Observed event rate"
  )
```

You bin predictions into deciles and compare: mean predicted probability vs observed event rate

The points lie fairly close to the diagonal in your plot, which supports that the risk score is not wildly miscalibrated.\

**How to present:** When the model says 20% risk, historically about \~20% of those days are truly high-AQI.

#### 5.2 “Wildfire-driven risk” diagnostic

Operationally useful check: compare risk with and without wildfire inputs.

```{r}
# Create two prediction datasets: with true WSEI vs WSEI set to 0
new_with <- df_m
new_wo   <- df_m %>% mutate(across(contains("wsei_hfi_k"), ~ 0))

p_with <- fitted(fit_rq3_90, newdata = new_with, scale = "response")[, "Estimate"]
p_wo   <- fitted(fit_rq3_90, newdata = new_wo, scale = "response")[, "Estimate"]

delta <- tibble(delta_risk = p_with - p_wo)

ggplot(delta, aes(delta_risk)) +
  geom_histogram(bins = 50) +
  coord_cartesian(xlim = c(-0.12, 0.12), ylim = c(0, 8000)) +
  labs(
    title = "Marginal risk uplift from wildfire features (posterior mean)",
    x = "P(high AQI | WSEI) - P(high AQI | WSEI=0)",
    y = "Count"
  )
```

Most mass is near 0, meaning **wildfire features don’t change risk much on most days**, which is expected (most days aren’t smoke-impacted).\

The important part is the **right tail**: when wildfire exposure is high, `delta_risk` should become meaningfully positive — that’s your “smoke episode” operational signal.

Wildfire features don’t trigger false alarms on normal days, but they materially increase risk on smoke-affected days.
